---
title: "Project1"
author: "Kelsey Garcia kg29435"
date: "10/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
suicide_total <- read_csv("/stor/home/kg29435/project1/suicide_total_deaths.csv")
internet_users <- read_csv("/stor/home/kg29435/project1/internet_users.csv")
happy_whr <- read_csv("/stor/home/kg29435/project1/hapiscore_whr.csv")
```
## 0. Introduction 

##### Now that we're in a world highly obsessed with the internet and overall perceived happiness, as well as in the midst of one of the greatest suicide rates in history, I thought it would be interesting to study actual data over a 10-year period on all three of these topics. Three datasets were used, all from gapminder.org where each individual dataset contained a different set of information concerning the three variables of internet use, happiness score, and total suicides. The internet users data represents the percentage of the population in each country in a given year that uses the internet. The happiness score data was collected by questionnaire's and manipulated to represent the percentage of happiness the reporting individuals felt. These results were averaged over each country and over each year from 2005 to 2015. Lastly, the total suicides data represented the number of deaths caused by self-inflicted injury for each country across each of the aforementioned years. Prior to actually performing the analysis, I wanted to see if there was a negative correlation with internet use and happiness score, meaning that countries that the internet more tended to be less happy. I also expected to see a positive correlation between happiness score and total suicides meaning that generally happier countries tend to have fewer deaths by suicide. However, it's important to note that this data does not analyze many confounding variables such as quality of life in each country and individual causes of mental illness.


## 1. Tidying: Rearranging Wide/Long 

##### Each dataset was pivoted longer prior to merging so that the values for the three variables anlyzed were each given their own column and the year was also given its own column.

#### *Tidying the 'internet_users' dataset*
```{r}
internet_users <- select(internet_users, -('1959':'2004'),-('2016':'2018'))
head(internet_users)
internet_users_long<-internet_users %>% 
                      pivot_longer(2:12, names_to="year", values_to="internet_users")
head(internet_users_long)
```

#### *Tidying the 'happy_whr'dataset*
```{r}
happy_whr <- select(happy_whr, -('2004'),-('2016':'2018'))
head(happy_whr)
happy_whr_long <- happy_whr %>% 
                  pivot_longer(2:12, names_to="year", values_to="happiness_score")
head(happy_whr_long)
```

#### *Tidying the 'suicide_total' score*
```{r}
suicide_total <- select(suicide_total,-('1989':'2004'))
head(suicide_total)
suicide_total_long <- suicide_total %>%
                      pivot_longer(2:12, names_to="year", values_to="total_suicides")
head(suicide_total_long)

## changing the "k"'s in total suicide values (ex/ 41k) to numeric values
suicide_total_long <- suicide_total_long %>% mutate(total_suicides=str_replace(total_suicides, "k","000")) %>% mutate(total_suicides=str_replace(total_suicides,"\\.","")) %>%  mutate(total_suicides= as.numeric(total_suicides))
suicide_total_long %>% filter(country=="United States") %>% head()
```

## 2. Joining/Merging

### Relevant data information prior to merging

#### *The distinct country IDs for 'internet_users_long'*

##### The 'internet_users_long' dataset has 51 countries distinct from the 'happy_whr_long' and 20 countries in its data distinct from 'suicide_total_long'. Many of the countries unique to the 'internet_users_long' dataset are located amongst the Pacific and the Carribbean islands. 
    
```{r}
## number of unique IDs from 'happy_whr_long'
anti_join(internet_users_long, happy_whr_long, by=c("country","year")) %>% summarize(n_distinct(country))
## what those unique IDs are compared to 'happy_whr_long'
anti_join(internet_users_long, happy_whr_long, by=c("country","year"))
## number of unique IDs form 'suicide_total_long'
anti_join(internet_users_long, suicide_total_long, by=c("country","year")) %>% summarize(n_distinct(country))
## what those unique IDs are compared to 'suicide_total_long'
anti_join(internet_users_long, suicide_total_long, by=c("country","year"))
```

#### *The distinct country IDs for 'happy_whr_long'*

##### The 'happy_whr_long' dataset has 1 country distinct from the 'internet_users_long' dataset and 1 country in its data distinct from 'suicide_total_long'. 'happy_whr_long' has Taiwan included in its data when 'internet_users_long' doesn't and Hong Kong, China in its data when 'suicide_total_long' doesn't.

```{r}
## number of unique IDs from 'internet_users_long'
anti_join(happy_whr_long, internet_users_long, by=c("country","year")) %>% summarize(n_distinct(country))
## what those unique IDs compared to 'internet_users_long'
anti_join(happy_whr_long, internet_users_long, by=c("country","year"))
## number of unique IDs from 'suicide_total_long'
anti_join(happy_whr_long, suicide_total_long, by=c("country","year")) %>% summarize(n_distinct(country))
## what those unique IDs are compared to 'suicide_total_long'
anti_join(happy_whr_long, suicide_total_long, by=c("country","year"))
```

#### *The distinct country IDs for 'suicide_total_long'*

##### The 'suicide_total_long' dataset has 1 country distinct from the 'internet_users_long' dataset and 32 countries in its data distinct from 'happy_whr_long'. 'suicide_total_long' also has Taiwan as a unique country that the 'internet_users_long' dataset doesn't contain. Again, many of the countries unique to the 'suicide_total_long' dataset not found in the 'happy_whr_long' dataset are located amongst the Pacific and the Carribbean islands. 

```{r}
## number of unique IDs from 'internet_user_long'
anti_join(suicide_total_long, internet_users_long, by=c("country","year")) %>% summarize(n_distinct(country))
## what those unique IDs compared to 'internet_users_long'
anti_join(suicide_total_long, internet_users_long, by=c("country","year"))
## number of unique IDs from 'happy_whr_long'
anti_join(suicide_total_long, happy_whr_long, by=c("country","year")) %>% summarize(n_distinct(country))
## what those unique IDs are compared to 'happy_whr_long'
anti_join(suicide_total_long, happy_whr_long, by=c("country","year"))
```

### Joining 'internet_users_long','happy_whr_long', and 'suicide_total_long' together for data analysis

##### Based on the information on unique country IDs for each dataset it will be most effective to perform two inner joins to only obtain and analyze countries that have information for all three of the different variables in question. The inner join is effective for that because it only includes observations that are shared by any two datasets being joined. Each dataset was joined to another by their shared country and year variables. There were 161 country IDs in common between all three datasets.

##### A full join was performed to calculate the number of observations lost from performing an inner join versus a full join which was found to be 2,376 country & year observations in total(4,147 observations in the full join and 1,771 observations in the inner join). The 'internet_users_long' dataset originally had 2,343 observations, 'happy_whr_long' had 1,793 observations, and 'suicide_total_long' had 2,134 observatoins.
```{r}
# stepwise joining of the three datasets
partialjoin <- inner_join(internet_users_long, happy_whr_long, by=c("country","year"))

completejoin <- inner_join(partialjoin, suicide_total_long,by=c("country","year"))

head(completejoin)

# total number of observations/rows - \1,771 observations\
str(completejoin)

# number of id variables in common between all three datasets - \161 countries\
completejoin %>% summarize(n_distinct(country)) 

# what country IDs are shared by all three datasets
completejoin %>% count(country)

# number of observations lost when removing the unique country IDs associated with the various data sets
partialfjoin <- full_join(internet_users_long, happy_whr_long, by=c("country"="year"))

full_join(partialfjoin, suicide_total_long, by=c("country","year")) # gives 4,147 observations
4147-1771 # a loss of 2,376 observations
```

## 3. Wrangling 

#### *Creating a new variable of the percent change in internet use in the US from one year to the next*
##### The percent change in internet use from one year to the next changed anywhere 1.3% from 2006 to 2007 to 14.6% from 2014 to 2015. The internet use values in the United States have not decreased since 2012.
```{r}
completejoin %>% 
  filter(country=="United States") %>% 
  mutate(pct_change_internet_use = (internet_users-lag(internet_users)) / lag(internet_users)) 
```

#### *Summary statistics on categorical variables*
##### Each of the 161 country IDs was observed 11 times (from the years 2005-2015). Each year was observed 161 times, once for each country.

```{r}
## obtaining counts of observations for each country ID
completejoin %>% group_by(country) %>% summarize(n())
## obtaining counts of observations in each year
completejoin %>% group_by(year) %>% summarize(n())
## obtaining counts of observations of each combination of country ID and year
completejoin %>% group_by(country,year) %>% summarize(n())
```


#### *Overall (ungrouped) summary statistics for numeric variables*
##### The average percent of the population that uses internet was found to be about 35% with a range from 0.182% to 98.2% of a country's population using the internet. The average happiness score amongst all countries was 54.02. In other words, people tend to rate their happiness as being about the middle of the scale, not too poor but not too great. The mean of the total suicides in any country was 15,150 but the median was 879 meaning that a smaller number of countries with really high suicide counts pulled up the mean. There were also 18 NA's found for the 'internet_users' data and 409 NA's found for the 'happiness_score' data.

```{r}
## overall summary statistics for 'internet_users' variable
completejoin %>% summarize(mean_internet_users=mean(internet_users,na.rm=T),
                           sd_internet_users=sd(internet_users,na.rm=T),
                           med_internet_users=median(internet_users,na.rm=T),
                            min_internet_users=min(internet_users,na.rm=T),
                              max_internet_users=max(internet_users,na.rm=T))

## overall summary statistics for 'happiness_score' variable
completejoin %>% summarize(mean_happiness_score=mean(happiness_score,na.rm=T),
                           sd_happiness_score=sd(happiness_score,na.rm=T),
                            med_happiness_score=median(happiness_score,na.rm=T),
                            min_happiness_score=min(happiness_score,na.rm=T),
                              max_happiness_score=max(happiness_score,na.rm=T))

## overall summary statistics for 'total_suicides' variable
completejoin %>% summarize(mean_total_suicides=mean(total_suicides,na.rm=T),
                           sd_total_suicides=sd(total_suicides,na.rm=T),
                            med_total_suicides=median(total_suicides,na.rm=T),
                            min_total_suicides=min(total_suicides,na.rm=T),
                              max_total_suicides=max(total_suicides,na.rm=T))

## number of NAs for each variable
completejoin %>% summarize_all(function(x)sum(is.na(x)))
```
#### *Summary statistics for numeric variables grouped by 'country'*
##### The data for each country across all 11 years observed was analyzed to find the mean, standard deviation, median, maximum, and minimum for all three variables. The happiness score data, specifically, was further analyzed and arranged to identify the top 10 happiest countries in the dataset.
```{r}
## obtaining 2 of 5 summary statistics & showing an example of pivoting wider
groupedsummarystats <- completejoin %>% group_by(country) %>%
                      summarize(mean_internet_users=mean(internet_users,na.rm=T),
                           sd_internet_users=sd(internet_users,na.rm=T),
                           med_internet_users=median(internet_users,na.rm=T),
                            min_internet_users=min(internet_users,na.rm=T),
                              max_internet_users=max(internet_users,na.rm=T),
                           mean_happiness_score=mean(happiness_score,na.rm=T),
                           sd_happiness_score=sd(happiness_score,na.rm=T),
                            med_happiness_score=median(happiness_score,na.rm=T),
                            min_happiness_score=min(happiness_score,na.rm=T),
                              max_happiness_score=max(happiness_score,na.rm=T),
                           mean_total_suicides=mean(total_suicides,na.rm=T),
                           sd_total_suicides=sd(total_suicides,na.rm=T),
                            med_total_suicides=median(total_suicides,na.rm=T),
                            min_total_suicides=min(total_suicides,na.rm=T),
                              max_total_suicides=max(total_suicides,na.rm=T))

                                                 
head(groupedsummarystats)

## pivoting wider
groupedsummarystats %>% 
  pivot_longer(2:16) %>% 
  separate(name, into=c("stat","variable1","variable2")) %>% 
  unite("variable", variable1, variable2, sep="_") %>% 
  pivot_wider(names_from = "variable", values_from = "value") %>% head()
```
#### Top 10 Happiest Countries Table
```{r}
top10happiness<- groupedsummarystats %>% 
                  select(1,7:11) %>% slice_max(mean_happiness_score,n=10) %>%
                  rename(Country=country,
                  Mean=mean_happiness_score,
                  StdDeviation=sd_happiness_score,
                  Median=med_happiness_score,
                  Maximum=max_happiness_score,
                  Minimum=min_happiness_score) %>% mutate(Rank=c(1:10))
library(knitr)

top10happiness %>% 
  kable(digits = 3, align = 'c')
```

#### *Correlation Values Between Each Variable*
##### Numeric values for the correlation between each variable were also calculated to test if the hypothesized relationships between the variables were true or not.
```{r}
variablecor <- completejoin %>% 
  select_if(is.numeric) %>% 
  cor(use="pair") %>% 
  as.data.frame %>% 
  rownames_to_column("var1") %>%
  pivot_longer(-1,names_to="var2",values_to="correlation")

variablecor
```


## 4. Visualizing

#### *Correlation Map of All Three Numeric Variables*
##### This map is showing the correlation relationships between each numeric variable in the dataset. The orange color represents a higher positive correlation value and the pink represents a smaller positive correlation value. There is no color for negative correlation values in the map because there were no negative correlations identified between the numeric variables in the dataset. This proves that the initial hypothesis that higher internet use resulted in lower happiness scores is not true. In fact, it would seem as though there is a pretty significant positive correlation between happiness score and internet use. However, this may be due to other confounding variables such as the development status of the country.
```{r}
variablecor %>% 
  ggplot(aes(var1,var2,fill=correlation)) +
  geom_tile() +
  scale_fill_gradient2(mid="pink",high="orange") + 
  geom_text(aes(label=round(correlation,4)),color = "black", size = 4) + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  coord_fixed() +
  labs(title="Correlation Matrix of the Numeric Variables")
```


#### *Histogram of Mean Happiness Score for Each Year from 2005 to 2015*
##### This graph shows how the mean happiness score for all of the countries changed across the years from 2005 to 2015. The mean happiness score seems to have stayed pretty consistent across the 11 years observed. The happiest year across the world in this time period seems to be 2009 while the least happy was 2005. 
```{r}
completejoin %>% ggplot(aes(year,happiness_score))+
                geom_histogram(stat="summary",fun=mean,color="black",fill="pink",alpha=.4,width=.8)+
                geom_errorbar(stat="summary",fun.data=mean_se,width=.8) + 
                theme(axis.text.x = element_text(angle=45, hjust=1)) + 
                scale_y_continuous(n.breaks=12)+
                labs(title= "Mean Happiness Score Across the Years")
```

#### *Scatterplot of Internet Users vs. Happiness Score*
##### This plot depicts the linear relationship between internet use and the happiness scores reported. This plot also depicts the overall positive correlation/linear relationship between these two variables showing that as the internet is used more, the happier the population as a whole seems to be. This linear relationship seems to flatten out somewhat around 50% population internet use and then becomes more strongly linear again (as representd by the geom_smooth trendline).
```{r}
completejoin %>% 
  ggplot(aes(internet_users,happiness_score)) + 
  geom_point(size=2) + 
  geom_jitter() + 
  geom_smooth(alpha=.5,color="orange") +
  labs(title="Internet Use vs. Reported Hapiness Scores")
```

